---
title: "Arrieta for Cy Young"
author: "Michael Griffiths"
date: "November 19, 2015"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 8
    fig_width: 12
    keep_md: yes
    toc: yes
---

<style type="text/css">
  .section {overflow-x:scroll;}
</style>

This is an effort to duplicate Greg Reida's [iPython notebook on Cy Young](https://github.com/gjreda/cy-young-NL-2015/blob/master/cy-young.ipynb) in R, for kicks and giggles.

To begin, let's load the Hadley-verse.

```{r Load_Libraries, warning=FALSE, comment=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

# For quick plot styles
library(ggthemes)
# Set to FiveThirtyEight default
theme_set(theme_fivethirtyeight())

# To customize output for HTML.
library(knitr)
library(pander)
knit_print.data.frame = function(x, options){ pander(x) }
knit_print.list = function(x, options){ pander(x) }
knit_print.matrix = function(x, options){ pander(x) }
knit_print.table = function(x, options){ pander(x) }

panderOptions('table.style', "rmarkdown")
panderOptions('table.split.table', Inf)
panderOptions('table.split.cells', Inf)
panderOptions('table.alignment.default', 'left')

```

## Load game log files

First, let's load in some of the initial game files.

```{r Load_Data}
# Convert date column to an actual date object. 
# Note that date comes in as either "Apr 11" or "Jul 12(1)"; we don't care about the brackets.
to_date <- function(datestring){
  date_no_brackets = gsub("\\(.+\\)", "", datestring)
  dates <- as.Date(sprintf("%s 2015", date_no_brackets), "%b %d %Y")
  return(dates)
}

# Convert innings to fraction.
# See @url(https://en.wikipedia.org/wiki/Innings_pitched)
innings_pitched <- function(ip){
  inning = floor(ip)
  partial_inning = ip %% 1
  return(inning + partial_inning / .3)
}

# Load three files into one dataframe, with a new column for the filename.
files = c("arrieta", "greinke", "kershaw")
data = list()
for(file in files){
  read_csv(sprintf("data/gamelogs/%s2015.csv", file)) %>%
    mutate(Date = to_date(Date), 
           player = file,
           IP = innings_pitched(IP)) ->
    data[[file]]
}
data = bind_rows(data)
head(data)
```

## Calculate conversions

Now that we have the initial data loaded, let's calculate the summary statistics we want to look at on a per-player basis.

```{r Calculate_Summary_Stats}

addStatisticsColumns <- function(dataframe){
  dataframe %>%
  mutate(rollingIP = cumsum(IP),
         IPGame = rollingIP / Rk,
         rollingER = cumsum(ER),
         rollingERA = rollingER / rollingIP * 9,
         rollingSO = cumsum(SO),
         strikeoutsPerIP = rollingSO / rollingIP,
         `K/9` = rollingSO / rollingIP * 9,
         strikeoutsPerBF = rollingSO / cumsum(BF),
         hitsPerIP = cumsum(H) / rollingIP,
         hitsPerAB = cumsum(H) / cumsum(AB),
         rollingWHIP = (cumsum(H) + cumsum(BB)) / rollingIP,
         # Opponents against
         `1B` = H - `2B` - `3B` - HR,
         AVG = cumsum(H) / cumsum(AB),
         OBP = (cumsum(H) + cumsum(BB) + cumsum(HBP)) / (cumsum(AB) + cumsum(BB) + cumsum(HBP) + cumsum(SF)),
         SLG = (cumsum(`1B`) + 2*cumsum(`2B`) + 3*cumsum(`3B`) + 4*cumsum(HR)) / cumsum(AB),
         OPS = OBP / SLG,
         # Rates
         BABIP = (cumsum(H) - cumsum(HR)) / (cumsum(AB) + cumsum(SO) + cumsum(HR) + cumsum(SF)),
         `HR%` = cumsum(HR) / cumsum(BF),
         `XBH%` = (cumsum(`2B`) + cumsum(`3B`) + cumsum(HR)) / cumsum(BF),
         `K%` = cumsum(SO) / cumsum(BF),
         `IP%` = (cumsum(AB) - cumsum(SO) - cumsum(HR) + cumsum(SF)) / cumsum(BF),
         `GB%` = cumsum(GB) / (cumsum(AB) - cumsum(SO) - cumsum(HR) + cumsum(SF))
         ) %>%
    return
}

data %>%
  group_by(player) %>%
  arrange(Date) %>%
  addStatisticsColumns %>%
  ungroup ->
  data
```

Since we've calculated all of out statistics (and there are a lot of them!) let's move on to the fun part - **graphs**:

```{r Initial_Comparison, fig.width=8, fig.height=8}
data %>%
  ggplot(aes(x=Date, y=`GB%`, group=player)) +
  geom_line(aes(colour=player), size=2) +
  ggtitle("Ground Ball Percentage")
```

We can see Arrieta continually improving, finally blasting past Kershaw.

## Arrieta's Second Half

Not sure what we do with this, but here how you do it - 

```{r}
allstarbreak = to_date('Jul 14')
data %>%
  filter(player == "arrieta", Date >= allstarbreak) %>%
  arrange(Date) %>%
  addStatisticsColumns ->
  arrieta2H
```

# Grid O' Stats

Now let's produce some more plots!

It's difficult to reproduce, exactly, the plot stlye in matplotlib. However, we can do an approximation of one of the rows very quickly:

```{r FirstPlot, fig.height=4, fig.width=12}

data %>%
  gather(vars, values, rollingERA, `K/9`, AVG, OBP, SLG) %>% 
  ggplot(aes(x=Date, group=player)) +
  geom_line(aes(y=values, colour=player), size=2) +
  facet_wrap(~vars, nrow=1, scales="free") 

```

Now let's do the same thing for more columns - 

```{r SecondPlot, fig.height=4, fig.width=12}

data %>%
  gather(vars, values, `IP%`, BABIP, `XBH%`, `HR%`, `K%`) %>% 
  ggplot(aes(x=Date, group=player)) +
  geom_line(aes(y=values, colour=player), size=2) +
  facet_wrap(~vars, nrow=1, scales="free") 

```

## Season Simulation

What if we replayed the season (sampling randomly from their performances)? Can we tell who is truly the ERA winner?

```{r Season_Simulation}
calculateSeasonStats <- function(season) {
  season %>%
    summarise(
      ERA = sum(ER) / sum(IP) * 9,
      SO = sum(SO),
      H = sum(H),
      `2B` = sum(`2B`),
      `3B` = sum(`3B`),
      HR = sum(HR),
      BB = sum(BB),
      HBP = sum(HBP),
      SF = sum(SF),
      AVG = H / sum(AB),
      OBP = (H + BB + HBP) / (sum(AB) + BB + HBP + SF),
      SLG = (H + 2*`2B` + 3*`3B` + 4*HR) / sum(AB),
      AB = sum(AB),
      BF = sum(BF),
      Pit = sum(Pit),
      Str = sum(Str),
      StL = sum(StL),
      StS = sum(StS)
    ) %>%
    return
}

sampleSeasonWithReplacement <- function(season, runs=1000){
  results = list()
  for(sample_idx in 1:runs) {
    season %>% 
      # By default, sample is 100% of original size.
      sample_frac(replace=T) %>%
      calculateSeasonStats ->
      results[[sample_idx]] 
  }
  df = bind_rows(results)
  return(df)
}

```

Now that we have some helper functions to sample, we can just execute against each player - 

```{r Run_Simulation}

data %>%
  group_by(player) %>%
  do(sampleSeasonWithReplacement(.)) %>%
  ungroup -> 
  simulation

```

That allows us to plot comparison histograms:

```{r Plot_Simulation, message=FALSE, fig.height=6}
simulation %>% 
  ggplot(aes(x=ERA)) + 
  geom_histogram() + 
  facet_grid(player ~ .) + 
  ggtitle("ERA by Player")
```

## Simulation Results

Now let's use some of the simulation results we have. 

We can take a look at the CDF, where you can see the same pattern as in the histograms but more clearly represented.

```{r CDF, fig.height=5, fig.width=4}
simulation %>%
  ggplot(aes(x=round(ERA, 2))) +
  stat_ecdf(aes(colour=player), size=2)
  
```

We can also look at histograms for most of the statistics - 

```{r message=F}
simulation %>%
  gather(vars, values, ERA, SO, AVG, OBP, SLG) %>%
  ggplot(aes(x=round(values, 2))) +
  geom_histogram(size=2) +
  facet_grid(player~vars, scales="free")
```

It's also interesting to look at the CDF for much the same view.

```{r fig.height=5}
simulation %>%
  gather(vars, values, ERA, SO, AVG, OBP, SLG) %>%
  ggplot(aes(x=round(values, 1))) +
  stat_ecdf(aes(colour=player), size=2) +
  facet_wrap(~vars, scales="free", nrow=1)
```

# Pitch f/x

## Load Pitch Data

Now let's take a look at pitching data.

```{r Load_Pitch_Data, message=FALSE, warning=FALSE}
files = c("arrieta", "greinke", "kershaw")
pitch_data = list()
for(file in files){
  read_csv(sprintf("data/pitchfx/%s.csv", file)) %>%
    mutate(player = file) ->
    pitch_data[[file]]
}
pitch_data = bind_rows(pitch_data)
head(pitch_data, n=2)
```

## Analyze Pitch Data

We can peek at pitch result - 

```{r Preview_PitchResult}
pitch_data %>% 
  group_by(player, pitch_result) %>% 
  summarise(value=n()) %>% 
  spread(player, value)
```

... and at bat results.

```{r Preview_AtBat}
pitch_data %>% 
  group_by(player, atbat_result) %>% 
  summarise(value=n()) %>% 
  spread(player, value)
```

We also want to add a few columns for strikes, etc.

```{r PitchData_AddColumns}
ball_vals = c('Ball', 'Ball In Dirt', 'Intent Ball', 'Hit By Pitch')
swing_and_miss = c('Swinging Strike', 'Swinging Strike (Blocked)', 'Missed Bunt')
hit_vals = c('Single', 'Double', 'Triple', 'Home Run')

# Add lookup table for at bat results -> bases equivalency.
at_bat_bases = data.frame(
  atbat_result=c("Single", "Double", "Triple", "Home Run"), 
  total_bases=1:4
)
  
pitch_data %>%
  left_join(at_bat_bases) %>%
  mutate(is_strike = ifelse(pitch_result %in% ball_vals, 0, 1),
         swing_and_miss = ifelse(pitch_result %in% swing_and_miss, 1, 0),
         is_hit = ifelse(pitch_result %in% hit_vals, 1, 0)
         ) ->
  pitch_data
```

Now let's see who hits harder!

```{r message=FALSE}
pitch_data %>%
  ggplot(aes(x=batted_ball_velocity)) +
  geom_histogram(aes(fill=player)) +
  facet_grid(. ~ player)
```

Let's also take a look for a few summary statistics.

```{r message=FALSE}
pitch_data %>% 
  filter(batted_ball_type != "") %>%
  ggplot(aes(x=batted_ball_velocity)) +
  geom_histogram() +
  facet_grid(player ~ batted_ball_type, scales="free")
```

Now, what about their pitching speed? We can see the realized distributoin - 

```{r warning=FALSE}
pitch_data %>% 
  filter(batted_ball_type != "") %>%
  ggplot(aes(x=batted_ball_velocity)) +
  geom_density(aes(fill=player, alpha=.2))
```

... but let's try to simulate the difference.

```{r}
set.seed(49)

simulatePitches = function(pitches, runs=1000){
  pitches %>%
    filter(!is.na(batted_ball_velocity)) %>%
    select(batted_ball_velocity) ->
    draws
  results = list()
  for(i in 1:runs){
    draws %>%
      sample_frac(replace=TRUE) %>%
      summarise(average_velocity = mean(batted_ball_velocity)) ->
      results[[i]]
  }
  results %>%
    bind_rows %>%
    return
}

# Run the simulation
pitch_data %>%
  group_by(player) %>%
  do(simulatePitches(.)) %>%
  ungroup ->
  pitches_simulated
head(pitches_simulated)
```

Now that we have some data on simulated pitches, let's take a look!

```{r}
pitches_simulated %>%
  ggplot(aes(x=average_velocity)) +
  geom_density(aes(fill=player, alpha=.3))
```

So, how do we know if they're significantly different?

The means for greinke and arrieta are certainly different - greinke is at ~88, whereas arrieta is at ~85.

```{r}
pitches_simulated %>%
  group_by(player) %>%
  summarise(velocity = mean(average_velocity))
```

We can also run a t-test

```{r}
v1 = pitches_simulated %>% filter(player=="arrieta") 
v2 = pitches_simulated %>% filter(player=="greinke")

# Run the test
t.test(v1$average_velocity, v2$average_velocity) ->
  t_arrieta_greinke
t_arrieta_greinke
```

We see there's a **very** low chance (p ~= `r t_arrieta_greinke$p.value`) that these came from the same underlying distribution. So: Greinke gets hit "harder" than the other two players, for whatever reason - perhaps his pitches are easier to hit head on.

#### What is the relationship between batted ball velocity and batting average against these three?

To replicate - count the number of non-null balls hit back.

```{r}
pitch_data %>%
  filter(!is.na(batted_ball_velocity)) %>%
  group_by(player) %>%
  summarise(`# of pitches batted` = n())
```

#### How difficult is each of their particular pitches to hit?

```{r}
pitch_data %>%
  group_by(player, pitch_type) %>%
  summarise(total = n()) %>%
  ungroup() %>%
  spread(player, total)
```

# Historical Cy Young Results

Now let's take a look at some historical results.

```{r warning=FALSE}
# Load data file and convert share % to a number.
read_csv("data/cyyoung/results.csv") %>%
  mutate(share = as.numeric(gsub("%", "", share)) / 100) ->
  results
```


```{r}
results %>%
  group_by(year, league) %>%
  # Note that we're using dense_rank here, which does *not* allow for gaps, but *does* allow for ties.
  # This gives different results than the Python code.
  # We could replicate with row_number() to an extent, but we'd need to order by a third column to get consistent results,
  # In the situation of ties.
  mutate(era_rank = dense_rank(desc(earned_run_avg)),
         wins_rank = dense_rank(desc(W)),
         # Add "winner" flag
         winner = ifelse(rank == 1, 1, 0)
         ) %>%
  ungroup ->
  results
```

Note that, since the way we're calculating rank is different, we get different results from the iPython notebook - 

```{r CalculateWinsRank_Winner}

results %>% 
  group_by(wins_rank) %>% 
  summarise(sum(winner))

```


```{r Rankings}

results %>%
  group_by(wins_rank, era_rank) %>%
  summarise(total=sum(winner)) %>%
  ungroup %>%
  spread(era_rank, total)

```
